\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{category}


\begin{document}
\firstpage{1}

\subtitle{Application Note}

\title[pulseR package]{pulseR: Versatile computational analysis of RNA turnover from metabolic labeling experiments}
\author[Uvarovskii \textit{et~al}.]{
Uvarovskii Alexey\,$^{\text{\sfb 1,2}*}$,
Christoph Dieterich\,$^{\text{\sfb 1,2}*}$ }
\address{$^{\text{\sf 1}}$
Section of Bioinformatics and Systems Cardiology,
Klaus Tschira Institute for Integrative Computational Cardiology,
Department of Internal Medicine III,
University Hospital Heidelberg,   
Im Neuenheimer Feld 669,
69120 Heidelberg,
and 
$^{\text{\sf 2}}$
German Center for Cardiovascular Research (DZHK),
Im Neuenheimer Feld 669
69120 Heidelberg
}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{\textbf{Motivation:} 
Metabolic labelling of RNA is a well established and powerful method to estimate RNA synthesis and decay rates.
The pulseR R package simplifies the analysis of RNA-seq count data that emerges from corresponding pulse-chase experiments. \\
\textbf{Results:} 
The pulseR package provides a flexible interface and readily accommodates numerous different experimental designs.
To our knowledge, it is the first publicly available software solution that models count data with the more appropriate negative-binomial model.
Moreover, pulseR handles both, labelled and unlabelled, spike-ins in its workflow and accounts for potential labeling biases (e.g. number of uridine residues). \\
\textbf{Availability:} The pulseR package is freely
available at https://github.com/dieterich-lab/pulseR  under the
GPLv3.0 licence \\
\textbf{Contact:} \href{a.uvarovskii@uni-heidelberg.de}{a.uvarovskii@uni-heidelberg.de} and \href{christoph.dieterich@uni-heidelberg.de}{christoph.dieterich@uni-heidelberg.de}\\
\textbf{Supplementary information:} Supplementary data are available \href{https://github.com/dieterich-lab/pulseR-paper/blob/master/R/supplementary.pdf}{https://goo.gl/xmonsg}}
\maketitle

\section*{Introduction}
\input{intro}
%\vspace*{-.5cm}
\section*{Implementation}
\input{appnote.tex}
\vspace*{-.5cm}
\section*{Discussion}
\input{discussion.tex}
\subsection*{Acknowledgements}
AU and CD would like to thank all members from the Dieterich Lab for their great input.
Special thanks go to Isabel Naarman-de Vries for all experimental support and insights.
\paragraph{Funding\textcolon} 
The work of AU and CM was kindly supported 
by the Klaus Tschira Stiftung gGmbH and the German Center for Cardiovascular Research (DZHK).
\vspace*{-0.6cm}
\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrvnat}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
\bibliography{ref.bib}


\end{document}
